---
title: "Customer Deduplication"
author: "Lebintiti Kobe"
format: html
editor_options: 
  chunk_output_type: console
---

# Project Brief

------------------------------------------------------------------------

> You have been hired by Ebuy Emporium, a new e-commerce startup. They have been up and running for a month and have had unexpected success.
>
> They are starting to have an active interest in their customer base :
>
> Who are their customers? What do they buy? What drives their purchasing behavior ?
>
> Before they do any serious analysis, however, they need to be able to count their customers, which is proving more difficult than anticipated.
>
> One problem is there are multiple sources of customer data , which are:
>
> -   The e-commerce platform’s customer database, where customer details are recorded when they sign up for an account online. This is where most of the customers’ details should be found
>
> -   The in-house CRM system, where customer details are recorded when they make a purchase over the phone or are otherwise onboarded as customers (except as a result of purchasing online with a registered account) .
>
> -   The raw transaction data, which we will hereafter refer to as “purchases” or “sales,” and which also contains purchases made “as a guest,” meaning customer records are not explicitly created at the time of purchase . The way that the data seperate guests from other customers is by giving other customers ids , while assigning no ids to guests . Also , only guest have customer details while registered / crm customers do not . 

------------------------------------------------------------------------

# Results Driven Approach 

## Understanding the Problem 

### 1. "count their customers" :

Here , we thinking we should come up with a dataset that allows our stakeholders to count all their customers . Meaning , we need a dataset that combines all customers that are spread across multiple databases . 
We thinking of our best estimate of a dataset in which each distinct customer is one record / row , one row per distinct customer 

### 2. kinds of customers 

-  Customers that signed up (Registered Online)
-  Customers in CRM system (CRM) customers
-  Customers that made purchases
-  Customers that made purchases as guests (Guest Customers)

### 3. Assumptions based of data 

-  Since customers can make purchases as guests , then there is a chance a customer made a purchase as a guest then later registered then made purchase as registered customer / later made purchase over phone . So we can expect duplicates in guests , customer database and crm database 

-  Also a registered customer may have purchased over phone , so we can expect duplicates between crm and customer database 

-  In online systems , most systems when a customer is registering they can type whatever as their name , also they allowed to make typos , and if same customer made purchase over phone , then entered details may differ because of typos and mispellings . 


## Plan 

1. Explore and validate all datasets . Make sure all assumptions about data hold . 

2. Decide on the common schema our dataset should have . Here since the goal is just to count customers nothing else , we thinking of just including columns we will use for deduplication , columns for data lineage (indicators for each data) . So customer details and id (dedplication) then indicator columns for data lineage . 

2. Seperate all possible customers into their own datasets , Add customer type indicator variables for data lineage , Remove exact duplications then recombine them into one full dataset .
(Here since guests are from purchase dataset , they will two indicators , purchases and guest)

3. Deduplicate our customers , meaning we keep one customer from customers with same details but different ids . But here we need a column to show all linked_ids (dropped customers ids) so when other people they can verify that we indeed removed the right customers / for data lineage


# Validate Data 

## Purchase Data

Make sure all assumptions holds , there are no business logic violation e.t.c 

```{r}
options(scipen = 999)
require(here)
require(tidyverse)
require(janitor)
require(purr)
require(skimr)

purchases <- read.csv(here("data/purchases.csv")) %>% as_tibble()
glimpse(purchases)
```

### Basic Cleaning

Make sure character variables are in a consistent format , dates are datest , e.t.c

```{r}
require(lubridate)

purchases <- purchases %>%
  mutate(event_time = as_datetime(event_time),
         across(is.character,
                ~str_to_lower(.x)%>%
                  trimws()),
         across(is.character,
                ~ifelse(.x=="",NA_character_,.x)))
glimpse(purchases)
```

### Explore the data

```{r}
skim(purchases)
```

-  We have 70k+ records over 60k+ sessions . Makes senses , customers can purchase more than 1 item per session . 

Though our data has no missing product and category id : 

-  We have over 16k+ missing category_codes 

-  We have over 5K missing brands of items purchased

This suggests a possible database with these missing brands and category codes 

-  Prices range from 0.8c to 2.5K , no business violations here 

-  We have around 18k records with missing customer ids , which corresponds to purchases made by guests 
-  We have around 53K records with missing customer details which suggest purchases made by registered/crm customers

---

Validate assumption that guest purchases will always have missing id , while other purchases wont have customer details but ids , and that these cant be true at the same time , a record cant be a purchase made by guest and registered customers at the same time 

```{r}
purchases <- purchases %>%
  mutate(is_guest = is.na(customer_id))

purchases %>%
  filter(is_guest & !is.na(customer_id))
purchases %>%
  filter(!is_guest & is.na(customer_id))
```

It returns an empty dataset showing that we cant have registered customer purchase that is guest purchase at the same time 

```{r}
purchases %>%
  janitor::tabyl(is_guest)
```

> Also , 25% of our purchases are by guests , the rest are by registered customers 

---

Figure out approxiamate number of possible guests and registered customers that made purchases

```{r}
guests <- purchases %>%
  select(starts_with("guest"))%>%
  na.omit()%>%
  distinct()

registered <- purchases %>%
  select(customer_id)%>%
  na.omit()%>%
  distinct()


n_total <- count(guests)+count(registered)
n_guest <- count(guests)
n_reg <- count(registered)

n_total
(nrow(guests)/(nrow(guests)+nrow(registered)))*100
```

> We have around 33k distinct customers that made purchases , assuming we no duplicates . And out of these , 25% are guests , rest are registered customers 
>
So its safe to assume the upper bound of all customers will be aound 33K , the number can be less if there are duplicates , or more if there are customers that a registered but did not make purchases 


## CRM and Customer Data

```{r}
crm <- read.csv(here("data/crm_export.csv")) %>% as_tibble()
customer <- read.csv(here("data/customer_database.csv")) %>% as_tibble()

glimpse(crm)
glimpse(customer)
```

We see that CRM and Customer database have same customer details as for guest detail except , they have an extra AGE column . 

### Basic Cleaning

Make sure all character variables are in a consistent format e.t.c 

```{r}
crm <- crm %>%
  mutate(across(is.character,
                ~str_to_lower(.x)%>%
                  trimws()%>%
                  replace(.x=="",NA_character_)))

customer <- customer %>%
  mutate(across(is.character,
                ~str_to_lower(.x)%>%
                  trimws()%>%
                  replace(.x=="",NA_character_)))
```

### Basic Validation 

```{r}
skim(crm)
skim(customer)

summary(crm)
```

We have around 8K customers in CRM database
We have around 24K customers in Customer Database

---

Both CRM and Customer databases have customers with Age of Zero , which cant be possible , lets check these . Others are too old to be using online shopping , but its not impossible . 

```{r}
crm %>%
  filter(age==0)
crm %>%
  filter(age>90)

customer %>%
  filter(age==0)
customer %>%
  filter(age>90)
```

> 120 in CRM and 330 in Customer database . We need to check these , they may be system errors , or customers did not complete their profiles . 

---

Lets see how many duplicate customers are in CRM 

```{r}
crm %>%
  count()

crm %>%
  distinct()%>%
  count()

crm %>%
  count() - crm %>% 
              select(-customer_id)%>%
              distinct()%>%
              count()
```

So about 400 CRM customers have duplicate customer details but same ids , meaning this are potentially real duplicate people since age and postcode were factored 

```{r}
customer %>%
  count()

customer %>%
  distinct()%>%
  count()

customer %>%
  count()- customer %>%
  select(-customer_id)%>%
  distinct()%>%
  count()
```

So about 3.5K Customer database customers have duplicate details but same ids , meaning this are potentially real duplicate people since age and postcode were factored .

We could de-duplicate these when looking at individual dataset , but we will do it once all datasets are combined , is easier that way 

# Decide on the common schema our datasets should follow 

```{r}
purchases
crm
customer
```

We just need id and all customer details columns for deduplications , so we can figure if two customers are different or not .
Then we need indicator variables for each dataset (customer type) for data lineage purposes 

# Seperate each customer into their own dataset 

```{r}
guests_c <- purchases %>%
  select(starts_with("guest"))%>%
  na.omit()%>%
  distinct()
colnames(guests_c) <- colnames(crm %>% select(-age,-customer_id))

purchases_c <- purchases %>%
  filter(!is.na(customer_id))%>%
  select(customer_id)%>%
  distinct()

crm_c <- crm %>%
  distinct()

customer_c <- customer %>%
  distinct()
```

---

Add indicators for customers that made purchases for data linkage 

```{r}
guests_c <- guests_c %>%
  mutate(is_guest = T)

purchases_c <- purchases_c %>%
  mutate(in_purchase = T)
```

# Combine datasets 

## 1. Guest and Purchases back to one

```{r}
all_customers <- purchases_c %>%
  bind_rows(guests_c)%>%
  mutate(age=NA_real_)

all_customers %>%
  slice_sample(n=20)
```

---
Replace NULLs in is_guest and is_purchase details 

```{r}
all_customers <- all_customers %>%
  mutate(is_guest = replace_na(is_guest,F),
         in_purchase = replace_na(in_purchase,T))

all_customers %>%
  slice_sample(n=20)
```

## 2.Merge CRM with all_customers data

```{r}
all_customers <- all_customers %>%
  left_join(y=crm_c,
            by = "customer_id",
            suffix = c("_sales","_crm"))

all_customers %>%
  slice_sample(n=20)
```

```{r}
all_customers <- all_customers %>%
  mutate(in_crm_base = !is_guest & !is.na(surname_crm) | !is.na(age_crm))

all_customers %>%
  slice_sample(n=20)

all_customers %>%
  count(in_crm_base)

crm_c %>%
  count()
```

We managed to find 7K missing customer details in CRM for missing details in purchases data out of 7.8K CRM customers

---

Tidy up our merged dataset 

```{r}

all_customers <- all_customers %>%
  mutate(first_name_sales = ifelse(in_crm_base , first_name_crm ,first_name_sales),
         surname_sales = ifelse(in_crm_base , surname_crm , surname_sales),
         postcode_sales = ifelse(in_crm_base, postcode_crm , postcode_sales),
         age_sales = ifelse(in_crm_base , age_crm , age_sales))%>%
  select(-ends_with("_crm"))%>%
  rename("first_name" = first_name_sales,
         "surname" = surname_sales,
         "postcode" = postcode_sales,
         "age" = age_sales)

all_customers %>%
  filter(in_crm_base)
```

### 3. Non-Matched CRM customers 

Find customers in CRM which we not matched in purchases data 

```{r}
no_match_crm <- setdiff(unique(crm_c$customer_id),
        unique(all_customers$customer_id)) 

no_match_crm %>% length()
```

So we have 700 customers in CRM that did not make purchases 
---

Add these to our merged data

```{r}
all_customers <- all_customers %>%
  bind_rows(crm_c %>%
              filter(customer_id %in% no_match_crm))

all_customers <- all_customers %>%
  mutate(in_crm_base = replace_na(in_crm_base,T),
         in_purchase = replace_na(in_purchase,F),
         is_guest = replace_na(is_guest,F))

all_customers %>%
  select(starts_with("in_"),is_guest) %>%
  is.na()%>%
  any()
```



## 3.Merge Customer database with all_customers data

```{r}
all_customers <- all_customers %>%
  left_join(y=customer_c,
            by = "customer_id",
            suffix = c("_sales","_cust"))

all_customers %>%
  slice_sample(n=20)
```

```{r}
all_customers <- all_customers %>%
  mutate(in_customer_base = !is_guest & !is.na(surname_cust) | !is.na(age_cust))

all_customers %>%
  slice_sample(n=20)

all_customers %>%
  count(in_customer_base)

customer_c %>%
  count()
```

We managed to find 22K missing customer details in Customer database for missing details in purchases data out of 23K CRM customers

---

Tidy up our merged dataset 

```{r}

all_customers <- all_customers %>%
  mutate(first_name_sales = ifelse(in_customer_base, first_name_cust ,first_name_sales),
         surname_sales = ifelse(in_customer_base, surname_cust , surname_sales),
         postcode_sales = ifelse(in_customer_base, postcode_cust , postcode_sales),
         age_sales = ifelse(in_customer_base, age_cust , age_sales))%>%
  select(-ends_with("_cust"))%>%
  rename("first_name" = first_name_sales,
         "surname" = surname_sales,
         "postcode" = postcode_sales,
         "age" = age_sales)

all_customers %>%
  filter(in_customer_base)%>%
  select(-in_purchase,-in_crm_base)
```

### 3. Non-Matched Customer database customers 

Find customers in Customer database which we not matched in purchases data 

```{r}
no_match_cust <- setdiff(unique(customer_c$customer_id),
        unique(all_customers$customer_id)) 

no_match_cust %>% length()
```

So we have 1400 customers in CRM that did not make purchases 
---

Add these to our merged data

```{r}
all_customers <- all_customers %>%
  bind_rows(customer_c %>%
              filter(customer_id %in% no_match_cust))

all_customers <- all_customers %>%
  mutate(in_crm_base = replace_na(in_crm_base,F),
         in_purchase = replace_na(in_purchase,F),
         is_guest = replace_na(is_guest,F),
         in_customer_base = replace_na(in_customer_base,T))

all_customers %>%
  select(starts_with("in_"),is_guest) %>%
  is.na()%>%
  any()
```

## Find customers in purchases data but not in RM nor Customer databases 

```{r}
all_customers %>%
  filter(if_any(contains("name"),~is.na(.))) %>%
  count()
```

We have 1248 customers that made purchases but are neither in CRM nor Customer Databases . 
This suggest maybe an existance of some database with all these info . 

# Deduplication 

Now that we have our merged data , its time to do some de-duplication .
We have seen in CRM and Customer Database , we had customers with same details but with different id , we need to remove these duplicate customers.

Before we do this we need to decide which variable we going to use for deduplication ...
In CRM / Customer database , we decided that customers with same names , postcode and age are the same ,  which is fairly reasonable , but guests customers have do not have age variable . 

So if we want perform deduplication across all types of customers then we do make sure variables we use are available for all types of customers 
Here , we will have to use names and postcode in order to include all customers , but this will not be as accurate as when we included age column ...

Here , we assume two customers with the same names and postcodes are the same . Which is not necessarily true in real life , but its rare , so its not that bad of an assumption

Remember , that in order to be transparent and have clear data lineage , we need to link ids ...
But guests customers dont have these ids , so we create dummy ids for them ...
A simple way to do this is by assigning negative ids to all unique guest customers 

## Assign Dummy Ids to records with no ids 

```{r}
n_guest <- n_guest %>% pull(n)
dummy_ids <- seq(-1,-n_guest,-1)

all_customers <- all_customers %>%
  mutate(customer_id = ifelse(is_guest,dummy_ids,customer_id))

all_customers %>%
  slice_sample(n=20)
```

## Manual Exact Deduplication 

Here , we assume that customers with the same names and postcode but different ids are the same customer 


---

First find all duplicate records and create linked_ids for them and drop the duplicate with preserving linked_ids

```{r}
exact_dupes <- all_customers %>%
  na.omit()%>%
  group_by(first_name,surname,postcode)%>%
  filter(n()>1)%>%
  mutate(linked_ids = list(customer_id))%>%
  distinct(first_name,surname,postcode,.keep_all = T)%>%
  ungroup()%>%
  select(customer_id,linked_ids)

exact_dupes %>%
  count()
```

4.5K customer details have duplicates in our purchases data

---

Join the linked_ids with our all customers dataset

```{r}
all_customers <- all_customers %>%
  left_join(y=exact_dupes,
            by = "customer_id")

all_customers%>%
  select(-starts_with("in_"))%>%
  slice_sample(n=20) %>%
  mutate(linked_ids = paste(linked_ids))
```

Make sure that we dont got redundant ids in linked_ids . Menaing a current customer_id should not be in the linked_ids list

```{r}
all_customers %>%
  mutate(dupe_ids = 
           map2(linked_ids,customer_id,
                ~setdiff(.x,.y)))%>%
  rowwise()%>%
  mutate(num_dupes = length(dupe_ids))%>%
  ungroup()%>%
  select(-starts_with("in_"),-postcode,-is_guest)%>%
  filter(num_dupes>2)%>%
  mutate(dupe_ids = paste(dupe_ids))
  
all_customers <- all_customers %>%
  mutate(dupe_ids = 
           map2(linked_ids,customer_id,
                ~setdiff(.x,.y)))%>%
  rowwise()%>%
  mutate(num_dupes = length(dupe_ids))%>%
  ungroup()

```

---

Drop duplicate in our all customers dataset

> distinct() find all distinct values including NA values , if we apply it to our dataset , it will only keep one out of all customers with no customer details , so we seperate these customers first then apply distinct , then rejoin these customers 

```{r}
na_customers <- all_customers %>%
  filter(if_any(c(first_name,surname,postcode),
                  ~is.na(.)))%>%
  distinct(customer_id ,.keep_all = T)%>%
  select(-linked_ids)

all_customers <- all_customers %>%
  distinct(first_name,surname,postcode,.keep_all = T)%>%
  select(-linked_ids)

all_customers <- all_customers %>%
  bind_rows(na_customers)%>%
  distinct()

all_customers %>%
  count()
```

We have a total of 28.6K unique customers after removing their exact duplicates 

---

# Advanced De-duplication 

Instead of just removing exact duplicates , we can take a step further . 
We can use method that detect if two entities are the same , that takes into consideration typos , mispellings , abbreviations e.t.c 

There are package out there to help us achieve this . 
In R im going to use reclin2 packages , reclin - record linkage 


## Now basic steps of record linkage 

1. Blocking (create candidate pair , blocking by some variable)
2. Comparison rules 
3. Similarity Scores Computation (use EM model for this)
4. Picking Threshold
5. Evaluating pairs for chosen threshold
6. Getting our final pairs

Also unlike distinct() , reclin2 does not match NA values , which is nice to know 

## Lets get it 

```{r}
require(reclin2)

## Create candidate pair with Blocking to compute pairs 
## only if they have same value for the block variable
## To speed up computation 

pairs <- pair_blocking(all_customers,
                       on = "postcode",
                       deduplication = T)
pairs


## Set up comparison rules 
## 

pairs <- compare_pairs(pairs = pairs, 
                       on = c("first_name",
                              "surname"),
                       comparators = list(cmp_jarowinkler(threshold = 1),
                                          cmp_jarowinkler(threshold = 1)))

pairs

## compute weights 
## 
pairs <- score_simple(pairs,variable = "scores",
                      on= c("first_name",
                              "surname"))
pairs

## select pairs for given threshold
## 
pairs <- select_threshold(pairs,"is_duplicated",score = "scores",threshold = 1.8)

pairs %>%
  filter(is_duplicated)

## GET ORIGINAL DATA WITH MATCHES INDICATED as clusters 
## 
results <- deduplicate_equivalence(pairs,
                                   "cluster",
                                   "is_duplicated") %>%
  as_tibble()

##
results %>%
  group_by(cluster)%>%
  filter(n()>1)%>%
  select(-is_guest,-in_purchase,-in_crm_base,-in_customer_base)%>%
  arrange(cluster)%>%
  print(n=100)%>%
  filter(num_dupes > 0)
```

Im happy with the results 

```{r}
final_data <- results %>%
  group_by(cluster)%>%
  mutate(other_ids = list(customer_id),
         other_ids = map2(
           other_ids , customer_id , 
           ~setdiff(.x,.y)))%>%
  rowwise()%>%
  mutate(n_dup = length(other_ids))%>%
  ungroup()%>%
  group_by(cluster)%>%
  slice_head(n=1)%>%
  arrange(cluster)%>%
  ungroup()

final_data %>%
  select(customer_id , dupe_ids , other_ids)


final_data <- final_data %>%
  mutate(
    all_linked_ids = map2(dupe_ids, other_ids, ~ {
      combined <- unique(c(.x, .y))
      if (length(combined) == 0) NULL else combined
    })
  )%>%
  select(-dupe_ids,-n_dup,-num_dupes,-cluster,-other_ids)


final_data <- final_data %>%
  relocate(-starts_with("in_"))%>%
  relocate(-is_guest)

final_data <- final_data %>%
  rowwise()%>%
  mutate(all_linked_ids = paste0("[",paste(all_linked_ids , collapse = ","),"]"))%>%
  mutate(all_linked_ids = ifelse(all_linked_ids == "[]",NA,all_linked_ids))%>%
  mutate(is_registered = ifelse(in_crm_base | in_customer_base , T , F))%>%
  ungroup()

results %>% count() - final_data %>% count() 
```

We managed to remove an Extra 882 duplicates using reclin2


With reclin2 we mane

---

Save data model to csv

```{r}
write.csv(x = final_data , here("data/data_model.csv"),row.names = F)
```

